tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-17 paper.tex
Updates from Daiki

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been identified~\citep{locke2015,ishigaki2022,
mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
This success has led to the collection of many population-scale Biobank
datasets~\citep{tanjo2021practical}.
This dense sampling and resulting vast volumes of data, however,
are presenting significant challenges to GWAS methology.
For example, how to correctly account for population stratification
remains an open question~\citep{uffelmann2021}.

@@

Simulation is a critical component of GWAS method development, and
generally consists of two steps: (1) generate a synthetic genetic
variation dataset; and (2) simulate some quantitative traits
based on this dataset.
(The two steps may sometimes be combined
in a single simulation, or we may use real data instead of simulations in
step (2), but the basic principles are still the same.)
The combined genetic variation and
simulated quantitative traits then represent the ``truth'' against
which association testing methods can be evaluated.
The first step, of generating realistic, large-scale genetic variation
datasets, is a well-studied problem with a mature and robust supporting
software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
kelleher2018efficient,haller2018tree,adrion2020,baumdicker2022efficient,anderson2023}
have been instrumental in making population genetic simulations
of the scale and realism
required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
These simulation methods are all based on the ``succinct tree
sequence'' ARG encoding~\citep{ralph2020,wong2023general},
using the open-source \texttt{tskit} library.

@@

There are therefore sophisticated methods available for simulating
ARGs at the population scale, but there is currently no
easy way to generate quantitative traits based on such an ARG.
Existing methods to simulate quantitative traits from a given set of
genetic sequences in standard formats~\citep[e.g.][]{meyer2018,fernandes2020}
are overwhelmed by the scale of the data. For example,
just parsing the 280TB of VCF for the French-Canadian
dataset mentioned above would take many days of CPU time,
making quantitative trait simulation by this approach unfeasible.
% TODO this needs another pass. We do need to address the Martin et
% al thing up front though, I think.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
used efficient \texttt{tskit} APIs to simulate quantitative
traits for 600,000 humans directly from a simulated ARG.
While the code provided by Martin et al.\ works at this
very large scale, it is tightly coupled to the details of their study
and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is therefore a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG
in \texttt{tskit}~\citep{ralph2020} format.
\texttt{Tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
Python data-science ecosytem, allowing users to efficiently analyse
large-scale data using familiar and ergonomic tools.

\section{Results}
\begin{figure}[t]%
\centering
\includegraphics[width=240pt]{figures/tree-illustration.pdf}
\caption{Example simulation of a phenotype at a site with ancestral state A and
two mutations. In this diploid example each of the three individuals
is assocated with two nodes (i.e., the individual with ID 0 corresponds
to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
with the null individual, $-1$. Here, the trait's causal allele is T
and effect size $\beta=0.05$. Each node in the tree has an associated
genetic value, and the overall genetic value for an individual is the
sum of their corresponding node values. The final phenotype
for each individual is the sum of the genetic value and
simulated environmental noise.
\label{fig:tree-illustration}}
\end{figure}

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
models, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
Each trait is associated with one or more causal sites (position
on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide). Because we are reasoning
about the inheritance of alleles at a site on the genome
we can consider just the local tree at that site and
the network structure of the ARG is not important~\citep{wong2023general}.
In Fig.~\ref{fig:tree-illustration} we show the local tree
relating three diploid individuals for a given causal site.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general the ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.

At a given site, the causal allele is associated with an
effect size $\beta$,
and every node in the tree that carries this allele has a
``genetic value'' of $\beta=0.05$. In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele, and therefore all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021}
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is normally distributed with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
% TODO I find this clearer than saying "additive genetic variance"
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.
% The \texttt{tstrait} API is modular, however, and other forms of environmental
% noise can be easily generated by the user, if they wish.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG library~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
In the simplest case, simulating a phenotype for an ARG requires
very little code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model, h2=0.3)
\end{lstlisting}
We first create \texttt{model} representing the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function, which also takes the trait model
and narrow sense heritability $h^2$ as parameters.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{pandas}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conventiently
processed by a host of Python data science tools.

@@

Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more
realistic data. For example, \texttt{tstrait} is used to simulate traits from
2.7 million individuals in the French Canadian dataset, which is the largest
known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to
simulate a trait with 1000 causal sites, making \texttt{tstrait} to be
practical to simulate quantitative traits from this largest ever simulated
dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is
also used to measure the computational performance.

>>>>

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been identified~\citep{locke2015,ishigaki2022,
mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
This success has led to the collection of many population-scale Biobank
datasets~\citep{tanjo2021practical}, 
but the sampling and resulting vast volumes of data
are presenting significant challenges to GWAS 
methology~\citep{uffelmann2021}.

@@

Simulation is a critical component of GWAS method development, and
generally consists of two steps: (1) generate a synthetic genetic
variation dataset; and (2) simulate quantitative traits
based on this dataset.
(The two steps may sometimes be combined
in a single simulation, or we may use real data instead of simulations in
step (2), but the basic principles are still the same.)
The combined genetic variation and
simulated quantitative traits then represent the ``truth'' against
which association testing methods can be evaluated.
The first step, of generating realistic, large-scale genetic variation
datasets, is a well-studied problem with a mature and robust supporting
software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
kelleher2018efficient,haller2018tree,adrion2020,baumdicker2022efficient,anderson2023}
have been instrumental in making population genetic simulations
of the scale and realism
required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
These simulation methods are all based on the ``succinct tree
sequence'' ARG encoding~\citep{wong2023general},
using the open-source \texttt{tskit} library~\citep{ralph2020}.

@@

There are therefore sophisticated methods available for simulating
ARGs at the population scale, but there is currently no
easy way to simulate quantitative traits based on such an ARG.
Existing methods to simulate quantitative traits from a given set of
genetic sequences in standard formats~\citep[e.g.][]{meyer2018,fernandes2020}
are overwhelmed by the scale of the data. For example,
just parsing the 280TB of VCF for the French-Canadian
dataset mentioned above would take many days of CPU time,
making quantitative trait simulation by this approach unfeasible.
% TODO this needs another pass. We do need to address the Martin et
% al thing up front though, I think.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
used efficient \texttt{tskit} APIs to simulate quantitative
traits for 600,000 humans directly from a simulated ARG.
While the code provided by Martin et al.\ works at this
very large scale, it is tightly coupled to the details of their study
and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG
in \texttt{tskit} format.
\texttt{Tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
Python data-science ecosystem~\citep{numpy}, 
allowing users to efficiently analyse
large-scale data using familiar and ergonomic tools.

\section{Results}
\begin{figure}[t]%
\centering
\includegraphics[width=240pt]{figures/tree-illustration.pdf}
\caption{Example simulation of a phenotype at a site with ancestral state A and
two mutations. In this diploid example each of the three individuals
is associated with two nodes (i.e., the individual with ID 0 corresponds
to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
with the null individual, $-1$. Here, the trait's causal allele is T
and effect size $\beta=0.05$. Each node in the tree has an associated
genetic value, and the overall genetic value for an individual is the
sum of their corresponding node values. The final phenotype
for each individual is the sum of the genetic value and
simulated environmental noise.
\label{fig:tree-illustration}}
\end{figure}

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
models~\citep{uffelmann2021}, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
Each trait is associated with one or more causal sites (position
on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide). Because we are reasoning
about the inheritance of alleles at a site on the genome,
we can consider just the local tree at that site and
the network structure of the ARG is not important~\citep{wong2023general}.
In Fig.~\ref{fig:tree-illustration} we show the local tree
relating three diploid individuals for a given causal site.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general the ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.

At a given site, the causal allele is associated with an
effect size $\beta$,
and every node in the tree that carries this allele has a
``genetic value'' of $\beta=0.05$. In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele, and therefore all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
% TODO I find this clearer than saying "additive genetic variance"
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.
% The \texttt{tstrait} API is modular, however, and other forms of environmental
% noise can be easily generated by the user, if they wish.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG library~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
In the simplest case, simulating a phenotype for an ARG requires
very little code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model, h2=0.3)
\end{lstlisting}
We first create \texttt{model} representing the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function, which also takes the trait model
and narrow sense heritability $h^2$ as parameters.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{pandas}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed by a host of Python data science tools.

@@

% TODO this is to be rewritten with just a single example from FC
% dataset and optionally also mentioning real data from 1000G (space
% permitting).
Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more
realistic data. For example, \texttt{tstrait} is used to simulate traits from
2.7 million individuals in the French Canadian dataset, which is the largest
known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to
simulate a trait with 1000 causal sites, making \texttt{tstrait} to be
practical to simulate quantitative traits from this largest ever simulated
dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is
also used to measure the computational performance.
